A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast

### 2022-02-24 analysis/paper/paper.Rmd
Minor edits

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)

@@

# Group dates that overlap with 99.7 % probability using the group_dates()
# function. All functions in this chunk where package isnt specified are defined 
# in 04script.R
date_groups <- group_dates(rcarb_sa, sitename)

@@

The negative values around 8000 BCE originate from the sites Løvås 1, 2 and 3. These are recently excavated, well-dated sites situated in a relatively undistributed area of the landscape. Thus, while there would be a danger of circularity of having archaeological sites inform a reconstruction RSL-change, and in turn use these to evaluate the degree of shore-bound settlement, the sites do clearly represent a upper limit for the sea-level, as they would not have been in use when located under water. Furthermore, as the Løvås sites are located at elevations from where the local displacement curve is interpolated between isololation basins at different elevation, it would seem that the sites represent a case where the archaeological material indicates discrepancies in the geological reconstruction of shoreline displacement in the area.

@@

# Shoreline dating the pioneer sites from Brunlanes

Not least following from the fact that very few pre-Boreal ^14^C-dates associated with anthropogenic activity have been achieved in Norway, the shoreline dating of the earliest sites is essential for understanding the pioneer settlement and the initial colonisation of the Scandinavian peninsula [e.g. @bang-andersen2012; @breivik2014; @glorstad2016]. The shoreline dated pre-Boreal sites from the Brunlanes-project are among the earliest known sites in Norway [@jaksland2014a; @jaksland2012; @jaksland2012a]. These have a distinct Early Mesolithic artefact inventory and are situated in a steep area of the landscape where it would be difficult to envision any use of the sites after the sea retreated any significant distance from their location. In the original publication of the sites, @jaksland2014 provides a thorough discussion of shoreline dating in general, and as used for the dating of the Brunlanes sites specifically. A comparison of his results and the ones achieved using the above-outlined approach are given in Figure Figure \@ref(fig:brunlanes)A. These are largely in agreement. The sites are located more or less on the isobase of the displacement curve for Larvik, and negligible adjustments were therefore needed. The small discrepancies between the results mainly follow from the fact that a slightly updated version of the displacement curve is applied here [cf. @sorenseninpress]. Jaksland's dates are given a flat 200 and 50 year uncertainty range starting from what he gives as the earliest possible date. The 200 year uncertainty range is given if the sites were to be considered in isolation, while the argument for the uncertainty range of only 50 years is based on the location of the sites relative to each other. Since they are located in such a constrained and steep area of the landscape, the difference in elevation between the sites is argued to establish their relative date and thus constrain the uncertainty ranges so that they don't overlap. This information is not integrated in the approach outlined here, but could justify reducing the the uncertainty ranges. The steepness of the curve does nonetheless provide very precise dates, although their accuracy is of course ultimately dependent on the veracity of the geological reconstruction. This example also highlights the fact that a consideration of the surrounding terrain and other sites can help in increasing the precision of the method if this can be used to exclude certain sea-levels as unlikely for when a site was in use. One approach to this could also be to assess the spatial implication of a proposed shoreline date by simulating the adjusted sea-levels, as is done for Pauler 1 in Figure \@ref(fig:brunlanes)B, followed for example by a visual evaluation of the topography or by measuring the steepness of the slope to the shoreline. Based on this, it could be possible to exclude certain elevations as unlikely for the position of the shoreline. Such approaches would make less of an impact in this setting, where the 95% probability range is already quite constrained, but could considerably improve the precision of the method in cases where the displacement curve is flatter (cf. Figure \@ref(fig:shoredate)). To further illustrate this point, excavated Stone Age sites in the region which have previously been dated with reference to relative sea-level change are subjected to the outlined approach below.

@@

In Figure \@ref(fig:redate)), excavated Stone Age sites within the study area where ^14^C-dates are not available or are not believed to date the main occupation of the sites are subjected to the method for shoreline dating outlined above. These are compared to the dates originally proposed in the excavation reports for the sites. This is an unfair exercise for a few reasons. First of all the dates provided in the reports are typically stated to be a very rough estimate, and are sometimes given as a point estimate with an implicitly or explicitly associated uncertainty. Secondly, seeing as these reports are from various dates in time, some are based on outdated data on RSL-change. Finally, they are often only meant to indicate a lower bound for when the sites could have been in use. I do, however, think that the results indicate that shoreline dating, in general, has been applied with a fairly reasonable degree of success, if these dates have been interpreted and informed research in an approximate manner. The results from applying the method devised here again highlights the spatial and temporal contingency of the method. In some cases the method provides a very precise date range and in others it offers little more than a *terminus post quem*.  Furthermore, a few of the date ranges extend well beyond the Stone Age and could be severely and securely constrained with only cursory reference to typology.

@@

# Discussion

When no other information is available it can be recommended to apply the method for shoreline dating outlined above, which involves both taking the distance between sites and the isobase of displacement curves into consideration, and implementing formula x to account for the distance between the sites and the shoreline. The precision of this method is, as shown above, both geographically and temporally contingent, where older sites situated towards the north-east will get a more precise date than younger sites located towards the south-west. This approach will be useful either when shoreline dating a large number of sites where no other information than site elevation is taken into account [e.g. @breivik2014; @solheim2018; @fossum2020; @nielsen2021b], which is typically done using a uniform range of uncertainty, likely overestimating the precision of the method in most cases and possibly underestimating it in some.

@@

Some sources of likely variation and uncertainty have not been considered here. First of all the DTM has only been corrected for major modern disturbances. This means that erosion, although likely not that prevalent, has not been taken into account. Secondly, the DTM has a vertical error [@lewis2021] ... Thirdly, the tidal range has not been considered...  Fourthly, the exposure of the sites to wave action is also likely to have been of concern [@roalkvam2020], and will presumably have implications for how close to the shoreline people settled [@blankholm2020]. And finally, the nature of the visits to the sites will presumably also have implications for how exposed a site. These dimensions will have implications for the shore-line dating for certain subsets of the sites and could be future avenues through which one could further improve the method. These dimensions will also be imprtant for more substantial considerations of the behavioural relation between sites and the shoreline.

# Conclusion

The most immediate contribution of this paper is a confirmation of previous research into the relation between coastal Norwegian Stone Age sites and the prehistoric shoreline, which has largely been verified. Furthermore, the typical application of shoreline dating as an approximate method where date ranges are often for example rounded off to 200-year intervals, means that most previous research based on the method is given support by the findings reported here. However, the findings also suggest that simply using the method as a terminus ante quem for the occupation of the sites is, at least in my view, overly conservative and would stand to unnecessarily loose important chronological information. Attempting to move these considerations further through an explicit quantification has here resulted in a concrete recommendation of applying an offset of 0-15m when shoreline dating sites where no other information is available.

@@

>>>>

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  fig.align = 'center',
  dpi = 300
)

@@

# Group dates that overlap with 99.7 % probability using the group_dates()
# function. (All functions in this chunk where package isn't specified are 
# defined in 04script.R).
date_groups <- group_dates(rcarb_sa, sitename)

@@

The negative values around 8000 BCE originate from the sites Løvås 1, 2 and 3. These are recently excavated, well-dated sites situated in a relatively undisturbed area of the landscape. Thus, while there would be a danger of circularity of having archaeological sites inform a reconstruction RSL-change, and in turn use these to evaluate the degree of shore-bound settlement, the sites do clearly represent a upper limit for the sea-level, as they would not have been in use when located under water. It could therefore seem that the Løvås sites represent a case where the archaeological material indicates discrepancies in the geological reconstruction of shoreline displacement in the area.

@@

Not least following from the fact that very few pre-Boreal ^14^C-dates associated with anthropogenic activity have been achieved in Norway, the shoreline dating of the earliest sites is essential for understanding the pioneer settlement and the initial colonisation of the Scandinavian peninsula [e.g. @bang-andersen2012; @breivik2014; @glorstad2016]. The shoreline dated pre-Boreal sites from the Brunlanes-project are among the earliest known sites in Norway [@jaksland2014a; @jaksland2012; @jaksland2012a]. These have a distinct Early Mesolithic artefact inventory and are situated in a steep area of the landscape where it would be difficult to envision any use of the sites after the sea retreated any significant distance from their location. In the original publication of the sites, @jaksland2014 provides a thorough discussion of shoreline dating in general, and as used for the dating of the Brunlanes sites specifically. A comparison of his results and the ones achieved using the above-outlined approach are given in Figure Figure \@ref(fig:brunlanes)A. The small discrepancies between the results mainly follow from the fact that a slightly updated version of the displacement curve is applied here [cf. @sorenseninpress]. Jaksland's dates are given a flat 200 and 50 year uncertainty range starting from what he gives as the earliest possible date. The 200 year uncertainty range is given if the sites were to be considered in isolation, while the argument for the uncertainty range of only 50 years is based on the location of the sites relative to each other. Since they are located in such a constrained and steep area of the landscape, the difference in elevation between the sites is argued to establish their relative date and thus constrain the uncertainty ranges so that they don't overlap. This information is not integrated in the approach outlined here, but could justify reducing the the uncertainty ranges. The steepness of the curve does nonetheless provide very precise dates, although their accuracy is of course ultimately dependent on the veracity of the geological reconstruction. This example highlights the fact that a consideration of the surrounding terrain and other sites can help in increasing the precision of the method if this can be used to exclude certain sea-levels as unlikely for when a site was in use. One approach to this could also be to assess the spatial implication of a proposed shoreline date by simulating the adjusted sea-levels, as is done for Pauler 1 in Figure \@ref(fig:brunlanes)B, followed for example by a visual evaluation of the topography or by evaluating the steepness of the slope to the shoreline. Based on this, it could be possible to exclude certain elevations as unlikely for the position of the shoreline for when the site was in use. Such approaches would make less of an impact in this setting, where the 95% probability range is already quite constrained, but could considerably improve the precision of the method in cases where the displacement curve is flatter (cf. Figure \@ref(fig:shoredate)).

@@

In Figure \@ref(fig:redate)), excavated Stone Age sites within the study area where ^14^C-dates are not available or are not believed to date the main occupation of the sites are also subjected to the method for shoreline dating outlined above. These are compared to the dates originally proposed in the excavation reports for the sites. To avoid issues with the DTM, the sites are dates based on the mean elevation of those provided for each site in the reports. This is an illustrative, but unfair exercise for a few reasons. First of all the dates provided in the reports are typically stated to be a very rough estimate, and are sometimes given as a point estimate with an implied or explicit, but undefined uncertainty range. Secondly, seeing as these reports are from various dates in time, many are based on now outdated data on RSL-change. Finally, they are often only meant to indicate a lower bound for when the sites could have been in use. I do, however, think that the results indicate that shoreline dating, in general, has been applied with a fairly reasonable degree of success, seeing as these dates have typically been interpreted and informed research in an approximate manner. The results from applying the method devised here again highlights the spatial and temporal contingency of the method. In some cases the method provides a very precise date range and in others it offers little more than a *terminus post quem*.  Furthermore, a few of the date ranges extend well beyond the Stone Age and could be severely and securely constrained with only cursory reference to typology. This thus points to another way that the dating method could be improved. In the above example from Brunlanes it was shown how this could conceivably be done by evaluating the geographical location of the sites. Another possibility is of course to draw on other temporal data, such as typological indicators or radiocarbon dates with which a shoreline date could be combined, for example within a Bayesian framework.

@@

# Concluding remarks

The most immediate contribution of this paper is what must be considered a confirmation of previous research into the relation between coastal Norwegian Stone Age sites and the prehistoric shoreline. This is indicated by the close relationship between sites and the shoreline up until the transition to the Neolithic at c. 3900 BCE after which a couple of sites become situated some distance from the shoreline, and the more decisive break at the transition to the Late Neolithic at c. 2400 BCE -- in clear agreement with the literature. Based on these findings, a refined method for the shoreline dating of Stone Age sites has been proposed. This involves both taking the distance between sites and the isobases of the displacement curves into consideration when dating the sites, and implementing formula x to account for the distance between the sites and the shoreline. When no other information is available, it can at present be recommended to use the empirically derived exponential decay ratios identified in Figure to characterise this relationship. However, the accuracy of the method can be improved by including more information, both with reference to the topographic location of the site and other temporal data such as typological indicators in the artefact inventories. The precision of the method is, as shown above, both geographically and temporally contingent due to the trajectory of RSL-change, where older sites situated towards the north-east in the study area will get a more accurate date than younger sites located towards the south-west. The impact of such additional information will therefore also vary.

Future investigations and radiocarbon dates from Stone Age sites in the region can not only be used to further test and adjust the findings reported here, but could also lay the foundations for refining the method. Given it's behavioural nature, it would for example seem likely that dimensions such as the nature and purpose of visits to the sites will have implications for how close to the shoreline they were located. Identifying such subsets of sites will thus potentially further refine the method, allowing it to be adjusted depending on the site being dated. Furthermore, other dimensions related to the topographic location of the sites could be similarly explored. This for example pertains to the exposure of sites to wave action, which is likely to have been of concern [@roalkvam2020], and which presumably has implications for how close to the shoreline people settled [@blankholm2020]. If such locational patterns can be discerned, this will also be useful for improving the shoreline dating of sites which have only been surveyed and where little information on the nature of the occupation of the site is available.

It is also worth highlithing some sources of likely variation and uncertainty have not been considered here. First of all the DTM has only been corrected for major modern disturbances. This means that erosion, although likely not that prevalent, has not been taken into account. Secondly, the DTM has a vertical error [cf. @lewis2021].

Thirdly, the tidal range has not been considered...   And finally, the nature of the visits to the sites will presumably also have implications for how exposed a site. These dimensions will have implications for the shore-line dating for certain subsets of the sites and could be future avenues through which one could further improve the method. These dimensions will also be imprtant for more substantial considerations of the behavioural relation between sites and the shoreline.

This approach will be useful either when shoreline dating a large number of sites where no other information than site elevation is taken into account [e.g. @breivik2014; @solheim2018; @fossum2020; @nielsen2021b], which is typically done using a uniform range of uncertainty, likely overestimating the precision of the method in most cases and possibly underestimating it in some.

@@

@@

 Furthermore, the typical application of shoreline dating as an approximate method where date ranges are often for example rounded off to 200-year intervals, means that most previous research based on the method is given support by the findings reported here. However, the findings also suggest that simply using the method as a terminus ante quem for the occupation of the sites is, at least in my view, overly conservative and would stand to unnecessarily loose important chronological information. Attempting to move these considerations further through an explicit quantification has here resulted in a concrete recommendation of applying an offset of 0-15m when shoreline dating sites where no other information is available.
